---
title: "Carbon Dioxide Avoidance"
author: "Jackie Culotta"
date: "2022-10-04"
output:
  html_document:
    df_print: paged
---

This Rmd file predicts dissolved CO~2~ (ppm) from instantaneous pH from water samples taken during avoidance trials. Thresholds for dissolved CO~2~ avoidance are estimated from pH using the resulting model.

```{r setup 1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()
```

# pH Trends in Tank

```{r CO2 map}
source("./LargeTank_CO2_Trend/CO2_Map.R")
pH_Trend_Both_Chambers
```

\
**The pH trend in both chambers.** The two chambers experienced relatively little water exchange across the connecting channel due to the circular chamber design and opposing currents.
However, water mixed uniformly within each chamber, as evidenced by similar pH values in the center of the chamber, immediately next to the exit, and at the gas stone.

```{r export CO2 map, include=FALSE, eval=FALSE}
ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig1_pH_Trend_Both_Chambers.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
pH_Trend_Both_Chambers
dev.off()

tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig1_pH_Trend_Both_Chambers.tif", width = 2100, height = 1200, units = "px", res = 300, compression = c("none"))
 pH_Trend_Both_Chambers
dev.off()
```

# CO2 avoidance

```{r load libraries, message=FALSE, echo=FALSE}
library('ggplot2')
library('ggpubr')
library('dplyr')
library('ggpmisc')
library('sjPlot')
library('knitr')
library('gdata') # keep function

keep(pH_Trend_Both_Chambers, sure = TRUE)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()
```

## Large Tank

```{r pH to ppm curve}
curve <- read.csv("allFish_pH_CO2_curve.csv") # All samples relating pH to pCO2 (ppm)
Exp1 <- read.csv("Exp1_LargeTank_CO2_Avoidance.csv") # Avoidance thresholds from last tank

# Create pH to CO2 ppm curve
Model_to_predict_pCO2_from_pH <- ggplot(curve, aes(x = pH, y = ppm)) +
  geom_point(size = 3) +
  stat_poly_line(formula = y ~ poly(x,2,raw=TRUE),
                 size = 2,se = FALSE,color = 'black') +
  stat_poly_eq(label.x = 5.0, formula = y ~ poly(x, 2, raw = TRUE),
               aes(label = paste(after_stat(eq.label),
                   after_stat(rr.label), sep = "*\", \"*"))) +
  scale_y_continuous(name = expression('pCO'[2]*' (ppm)')) +
  scale_x_continuous(name = "pH", breaks = seq(6,8.5,0.5)) + 
  theme_classic2()
Model_to_predict_pCO2_from_pH
```

```{r export CO2 curve figure, include=FALSE, eval=FALSE}
ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig2_Model_to_predict_pCO2_from_pH.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
Model_to_predict_pCO2_from_pH
dev.off()

tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig2_Model_to_predict_pCO2_from_pH.tif", width = 2100, height = 1200, units = "px", res = 300, compression = c("none"))
Model_to_predict_pCO2_from_pH
dev.off()
```


```{r pH to ppm model}
# Calculate 2nd order equation
fit2 <- lm(ppm~poly(pH,2,raw=TRUE), data = curve)
summary(fit2) # graph rounds coefficients in this equation

# Predict CO2 ppm at last avoidance pH
Exp1$pH <- Exp1$Last_Avoid_pH
Exp1$Last_Avoid_Estimated_CO2_ppm <- predict(fit2,newdata = Exp1)

# summarize results in large tank
Exp1_Results <- Exp1 %>% 
  filter_at(vars(X1st_Avoid_Measured_CO2_ppm, End_Trial_Measured_CO2_ppm), all_vars(!is.na(.))) %>% 
  group_by(Species) %>% 
  summarize(
  # FirstAvoid_Measured_ppm = mean(X1st_Avoid_Measured_CO2_ppm),
  # FirstAvoid_SE = sd(X1st_Avoid_Measured_CO2_ppm)/sqrt(length(X1st_Avoid_Measured_CO2_ppm)),
  Alk = mean(Alkalinity_ppm),
  Alk_SE = sd(Alkalinity_ppm)/sqrt(length(Alkalinity_ppm)),
  Temp = mean(Temp_C),
  Temp_SE = sd(Temp_C)/sqrt(length(Temp_C)),
   LastAvoid_Calculated = mean(Last_Avoid_Estimated_CO2_ppm, na.rm = TRUE),
   LastAvoid_SE = sd(Last_Avoid_Estimated_CO2_ppm)/sqrt(length(Last_Avoid_Estimated_CO2_ppm)),
   Measured = mean(End_Trial_Measured_CO2_ppm),
   Measured_SE = sd(End_Trial_Measured_CO2_ppm)/sqrt(length(End_Trial_Measured_CO2_ppm))
  )
Exp1_Results
```

```{r model summary as table, eval=FALSE, include=FALSE}
Format table to share values from 2nd order equation
library('stargazer')
stargazer(fit2,
         type = "html",
         out = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/Supplement/pH to CO2 Model Table.doc")
```

## Small Tank

```{r small tank data}
SmallExp1 <- read.csv("Exp1_SmallTank_CO2_Avoidance.csv")
SmallExp1$Temp_C <- as.numeric(SmallExp1$Temp_C)

# summarize results in small tank
SmallExp1_Results <- SmallExp1 %>% 
#  group_by(Species) %>% # not a high enough N for each speices
  summarize(
    Alk = mean(Alkalinity_Pre),
    Alk_SE = sd(Alkalinity_Pre)/sqrt(length(Alkalinity_Pre)),
    Temp = mean(Temp_C),
    Temp_SE = sd(Temp_C)/sqrt(length(Temp_C)), # unclear why this returns NA
    LastAvoid_Calculated = mean(CO2_Output_uAtm),
    LastAvoid_SE = sd(CO2_Output_uAtm)/sqrt(length(CO2_Output_uAtm)),
    Measured = mean(measured_pCO2_ppm, na.rm = TRUE),
    Measured_SE = sd(measured_pCO2_ppm, na.rm = TRUE)/sqrt(5)
  )
SmallExp1_Results
```

# Avoidance Thresholds

```{r publication table}
# CO2 Avoidance Table
SmallExp1_Results$Species <- c("Bigheaded")
SmallExp1_Results$Tank <- c("Small")
Exp1_Results$Tank <- c("Large")
BothExp1_Results <- full_join(SmallExp1_Results, Exp1_Results) # merge small and large tank results
BothExp1_Results <- select(BothExp1_Results, Tank, Species, everything()) # rearrange columns
kable(BothExp1_Results, 
      caption = "Carbon dioxide avoidance in both tanks",
      digits = 1)
```

```{r pull avoidance thresholds, include=FALSE}
attach(Exp1_Results)
# Calculate average avoidance threshold for all data
avoidAvgBoth <- round(mean(BothExp1_Results$LastAvoid_Calculated)) # option 2

# large tank
avoidBighead <- BothExp1_Results %>%  filter(Species == 'Bighead') %>%  select(LastAvoid_Calculated)
avoidSilver <- BothExp1_Results %>%  filter(Species == 'Silver') %>%  select(LastAvoid_Calculated)
avoidAvgLarge <- (avoidBighead + avoidSilver)/2
measureAvgLarge <- round(mean(Exp1_Results$Measured))

detach(Exp1_Results)
```

```{r Format table for publication, include=FALSE, eval=FALSE}
labels <- c("Tank", "Species", "TA (ppm)", "SE", "Temp (°C)", "SE","Estimated (ppm)","SE","Measured (ppm)","SE")
tab_df(BothExp1_Results,
       title = "Dissolved carbon dioxide avoidance and ambient water parameters",
       col.header = labels,
       footnote = "text",
       show.footnote = FALSE,
       encoding = "Windows-1252",
       digits = 1,
       file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/CO2 Avoidance Table.doc"
    #  user.viewer = TRUE
)
```

## CO2 avoidance boxplots figure

```{r Boxplot Figure, message=FALSE, warning=FALSE}
SmallExp1$Tank <- c("Small")
Exp1$Tank <- c("Large")

SmallExp1$Species <- gsub("S", "Bigheaded", fixed = TRUE, SmallExp1$Species) # grouping small tank across species
SmallExp1$Species <- gsub("BH", "Bigheaded", SmallExp1$Species)

BothExp1 <- full_join(Exp1, SmallExp1) %>% 
  mutate(Avoid = coalesce(Last_Avoid_Estimated_CO2_ppm,CO2_Output_uAtm))  %>% 
  select(Species,Avoid,Tank)

CO2_Avoidance_Boxplots <-  ggplot(BothExp1, aes(x = Species, y = Avoid, fill = Tank)) +
  stat_compare_means(label.x = 2.5) +
  geom_boxplot(position = position_dodge(1)) +
  geom_dotplot(binaxis = 'y',
               stackdir = 'center',
               position = position_dodge(1)) + # add dots
  labs(x = "Species", 
       y = expression("pCO"[2]*" (ppm) at Last Avoidance ")) +
  theme_classic2() + scale_fill_grey()
```

```{r output boxplot file, include=FALSE, eval=FALSE}
tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig5_CO2_Avoidance_Boxplots.tif", width = 1800, height = 1200, units = "px", res = 300, compression = c("none"))
CO2_Avoidance_Boxplots
dev.off()

ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig5_CO2_Avoidance_Boxplots.eps",
       dpi = 300, width = 6, height = 5, units = "in", limitsize = TRUE)
CO2_Avoidance_Boxplots
dev.off()

```

## pH Avoidance Thresholds

Calculate the pH at the avoidance threshold using the reverse equation from before: now ppm to pH

```{r Avoidance Thresholds}
fit <- lm(pH~poly(ppm,2,raw=TRUE), data = curve)

pH <- as.numeric(c("NA"))
ppm <- as.numeric(c(avoidAvgBoth))
threshold <- data.frame(pH, ppm)

threshold$pH <- predict(fit, newdata = threshold)
threshold$avoidAvgRound <- round(avoidAvgBoth, digits = 1)
threshold
```

```{r figure and kruscal test}
mod <- kruskal.test(Avoid ~ Species, data = BothExp1)
mod
```

```{r formating variables, include = FALSE}
# paragraph one: both tanks
chi2 <- format(round(as.numeric(mod$statistic), 2), nsmall = 2, big.mark = ",")
pval <- format(round(as.numeric(mod$p.value), 2), nsmall = 2, big.mark = ",")
avoidPPM <- format(round(as.numeric(avoidAvgBoth), 1), nsmall = 1, big.mark = ",")
pHThreshold <- format(round(as.numeric(threshold$pH), 2), nsmall = 2, big.mark = ",")

# paragraph two 
smallCO2 <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(LastAvoid_Calculated)), 1), nsmall = 1, big.mark = ",")
smallCO2_SE <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(LastAvoid_SE)), 1), nsmall = 1, big.mark = ",")
smallMeasured <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(Measured)), 1), nsmall = 1, big.mark = ",")
smallMeasured_SE <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(Measured_SE)), 1), nsmall = 1, big.mark = ",")

# paragraph three
largeBighead <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Bighead') %>% select(LastAvoid_Calculated)), 1), nsmall = 1, big.mark = ",")
largeBighead_SE <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Bighead') %>% select(LastAvoid_SE)), 1), nsmall = 1, big.mark = ",")
largeSilver <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Silver') %>% select(LastAvoid_Calculated)), 1), nsmall = 1, big.mark = ",")
largeSilver_SE <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Silver') %>% select(LastAvoid_SE)), 1), nsmall = 1, big.mark = ",")
largeBoth <- format(round(as.numeric(avoidAvgLarge), 1), nsmall = 1, big.mark = ",")
largeMeasured <- format(round(as.numeric(measureAvgLarge), 1), nsmall = 1, big.mark = ",")

largeMeasured_SE <- sd(Exp1$End_Trial_Measured_CO2_ppm, na.rm = TRUE)/sqrt(length(Exp1$End_Trial_Measured_CO2_ppm))
largeMeasured_SE <- format(round(as.numeric(largeMeasured_SE), 1), nsmall = 1, big.mark = ",")
largeMeasured_N <- as.numeric(length(Exp1$End_Trial_Measured_CO2_ppm))
```

# Results

```{r graph, echo=FALSE, message=FALSE}
CO2_Avoidance_Boxplots
```

\
**CO~2~ avoidance boxplots.** Bighead and silver carp avoided dissolved carbon dioxide at the same threshold level in the small and large tanks (`r mod$method`, Χ^2^ = `r chi2`, p = `r pval`).
Carp avoided pCO~2~ at `r avoidPPM` ppm.\

\
There is not a significant difference in dissolved CO~2~ avoidance thresholds between species or tank size (`r mod$method`, Χ^2^ = `r chi2`, p = `r pval`, Fig 1).
Across all trials in both tanks, bigheaded carp avoided pCO~2~ at `r avoidPPM` ppm.
Previous studies have found multiple species share similar CO~2~ avoidance thresholds (CITATION 17).
This threshold pCO~2~ level was used in CO~2~ conditioning treatments, by reaching an instantaneous pH of approximately `r pHThreshold` by the conditioning midpoint.

In the small shuttle tank, schools of five juvenile untrained bighead (n = 5) and silver (n = 5) carp exhibited CO~2~ avoidance at `r smallCO2` ± `r smallCO2_SE` ppm (n = 10, mean ± SE, Table 1) as estimated by instantaneous pH, TA, and temperature in CO2Calc.
Avoidance occurred at `r smallMeasured` ± `r smallMeasured_SE` ppm (n = 6) as measured from water samples by NDIR probe.
In the large shuttle tank, individual bighead carp (n = 10) exhibited their last CO~2~ avoidance at `r largeBighead` ± `r largeBighead_SE` ppm (Table 1).
Schools of five silver carp (n = 9) last avoided the increasing chamber at `r largeSilver` ± `r largeSilver_SE` ppm.
In the large tank, the last avoidance occurred at `r largeBoth` ppm CO~2~ on average for both species, as estimated by pH (S1).
The average pCO~2~ in the increasing chamber was measured at `r largeMeasured` ± `r largeMeasured_SE` ppm (n = `r largeMeasured_N`) three minutes after the last avoidance.
