---
title: "Carbon Dioxide Avoidance"
author: "Jackie Culotta"
date: "2022-10-04"
output:
  word_document: default
  html_document:
    df_print: paged
---

This Rmd file predicts dissolved CO~2~ (ppm) from instantaneous pH from water samples taken during avoidance trials.
Thresholds for dissolved CO~2~ avoidance are estimated from pH using the resulting model.

```{r setup 1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()
```

# pH Trends in Tank

```{r CO2 map, message=FALSE, warning=FALSE}
source("./CO2_Map.R")
pH_Trends_Both_Tanks
```

\
**Fig 1. A.** Water pH is plotted versus time (seconds) in the large and small shuttle tanks.
Time zero represents CO~2~ bubble initiation in the experimental chamber.
The dotted line indicates the approximate threshold for CO~2~ avoidance in bigheaded carps.

```{r export CO2 map, eval=FALSE, include=FALSE}
ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig1_pH_Trend_Both_Chambers.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
pH_Trend_Both_Chambers
dev.off()

tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig1_pH_Trend_Both_Chambers.tif", width = 2100, height = 1200, units = "px", res = 300, compression = c("none"))
 pH_Trend_Both_Chambers
dev.off()
```

# CO2 avoidance

```{r load libraries, echo=FALSE, message=FALSE, warning=FALSE}
library('ggplot2')
library('ggpubr')
library('dplyr')
library('ggpmisc')
library('sjPlot')
library('knitr')
library('gdata') # keep function

keep(pH_Trends_Both_Tanks, sure = TRUE)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# getwd()
```

## Large Tank

Create the figure for the model that predicts pCO~2~ (ppm) from pH.

```{r pH to ppm curve}
curve <- read.csv("allFish_pH_CO2_curve.csv") # All samples relating pH to pCO2 (ppm)

# Create pH to CO2 ppm curve
Model_to_predict_pCO2_from_pH <- ggplot(curve, aes(x = pH, y = ppm)) +
  geom_point(size = 3) +
  stat_poly_line(formula = y ~ poly(x,2,raw=TRUE),
                 size = 2,se = FALSE,color = 'black') +
  stat_poly_eq(label.x = 5.0, formula = y ~ poly(x, 2, raw = TRUE),
               aes(label = paste(after_stat(eq.label),
                   after_stat(rr.label), sep = "*\", \"*"))) +
  scale_y_continuous(name = expression('pCO'[2]*' (ppm)')) +
  scale_x_continuous(name = "pH", breaks = seq(6,8.5,0.5)) + 
  theme_classic2()
Model_to_predict_pCO2_from_pH
```

\
**Fig 1. B. Model to predict pCO~2~ from pH in the large tank.** Dissolved CO~2~ is plotted versus pH in the large tank.
Water samples were collected from the increasing pCO~2~ chamber after first exit and three minutes after the last exit.
The solid line represents the model described by the equation above (n = 72, F = 209, p \< 0.0001, adj R2 = 0.854).
Water pH was used as a proxy to determine CO~2~ concentrations according to the model.

```{r export figure 1, include=FALSE, eval=FALSE}
Fig1 <- ggarrange(pH_Trends_Both_Tanks, Model_to_predict_pCO2_from_pH,
                  labels = c("A","B"),
                  ncol = 1, nrow = 2,
                 # align = "v",
                  widths = c(1.5,1))
Fig1

ggsave(file = "./Figures/EPS/Fig1.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
Fig1
dev.off()

tiff("./Figures/TIF/Fig1.tif", width = (7*300), height = (6*300), units = "px", res = 300, compression = c("none"))
Fig1
dev.off()
```

Use the model to predict the pCO2 (ppm) that carp avoided, based on pH values.

```{r pH to ppm model}
Exp1 <- read.csv("Exp1_LargeTank_CO2_Avoidance.csv") # Avoidance thresholds from last tank

# Calculate 2nd order equation
fit2 <- lm(ppm~poly(pH,2,raw=TRUE), data = curve)
summary(fit2) # The figure above rounds coefficients in this equation

# Predict CO2 ppm at last avoidance pH
Exp1$pH <- Exp1$Last_Avoid_pH
Exp1$Last_Avoid_Estimated_CO2_ppm <- predict(fit2,newdata = Exp1)

# summarize results in large tank
Exp1_Results <- Exp1 %>% 
  filter_at(vars(X1st_Avoid_Measured_CO2_ppm, End_Trial_Measured_CO2_ppm), all_vars(!is.na(.))) %>% 
  group_by(Species) %>% 
  summarize(
  # FirstAvoid_Measured_ppm = mean(X1st_Avoid_Measured_CO2_ppm),
  # FirstAvoid_SE = sd(X1st_Avoid_Measured_CO2_ppm)/sqrt(length(X1st_Avoid_Measured_CO2_ppm)),
  Alk = mean(Alkalinity_ppm),
  Alk_SE = sd(Alkalinity_ppm)/sqrt(length(Alkalinity_ppm)),
  Temp = mean(Temp_C),
  Temp_SE = sd(Temp_C)/sqrt(length(Temp_C)),
   LastAvoid_Calculated = mean(Last_Avoid_Estimated_CO2_ppm, na.rm = TRUE),
   LastAvoid_SE = sd(Last_Avoid_Estimated_CO2_ppm)/sqrt(length(Last_Avoid_Estimated_CO2_ppm)),
   Measured = mean(End_Trial_Measured_CO2_ppm),
   Measured_SE = sd(End_Trial_Measured_CO2_ppm)/sqrt(length(End_Trial_Measured_CO2_ppm))
  )
Exp1_Results
```

```{r export model summary as table, eval=FALSE, include=FALSE}
# Format table to share values from 2nd order equation
library('stargazer')
stargazer(fit2,
         type = "html",
         out = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/Supplement/pH to CO2 Model Table.doc")
```

## Small Tank

Water parameters were different in the small tank compared to the large, so the same model cannot be used.
Instead, CO2Calc software was used to predict pCO~2~ from pH, temp, and TA.
Those results are summarized here.

```{r small tank data}
SmallExp1 <- read.csv("Exp1_SmallTank_CO2_Avoidance.csv")
SmallExp1$Temp_C <- as.numeric(SmallExp1$Temp_C)

# summarize results in small tank
SmallExp1_Results <- SmallExp1 %>% 
#  group_by(Species) %>% # not a high enough N for each species
  summarize(
    Alk = mean(Alkalinity_Pre),
    Alk_SE = sd(Alkalinity_Pre)/sqrt(length(Alkalinity_Pre)),
    Temp = mean(Temp_C),
    Temp_SE = sd(Temp_C)/sqrt(length(Temp_C)), # unclear why this returns NA
    LastAvoid_Calculated = mean(CO2_Output_uAtm),
    LastAvoid_SE = sd(CO2_Output_uAtm)/sqrt(length(CO2_Output_uAtm)),
    Measured = mean(measured_pCO2_ppm, na.rm = TRUE),
    Measured_SE = sd(measured_pCO2_ppm, na.rm = TRUE)/sqrt(5)
  )
SmallExp1_Results
```

# Avoidance Thresholds

Generate the table with water parameters, measured, and predicted pCO~2~ values at last exit from the increasing pCO~2~ chamber.

**Table 1. Dissolved carbon dioxide avoidance and ambient water parameters.** The pCO~2~ at last exit is estimated from instantaneous pH in both tanks.

```{r create co2 avoidance table}
# CO2 Avoidance Table
SmallExp1_Results$Species <- c("Bigheaded") # Bigheaded refers to both silver and bighead carps
SmallExp1_Results$Tank <- c("Small")
Exp1_Results$Tank <- c("Large")
BothExp1_Results <- full_join(SmallExp1_Results, Exp1_Results) # merge small and large tank results
BothExp1_Results <- select(BothExp1_Results, Tank, Species, everything()) # rearrange columns

CO2_Avoidance_Table <- kable(BothExp1_Results, 
     # caption = "Carbon dioxide avoidance and ambient water parameters in both tanks.",
      digits = 1)

CO2_Avoidance_Table
```

```{r pull avoidance thresholds for in-text render, include=FALSE}
attach(Exp1_Results)
# Calculate average avoidance threshold for all data
avoidAvgBoth <- round(mean(BothExp1_Results$LastAvoid_Calculated))

# large tank
avoidBighead <- BothExp1_Results %>%  filter(Species == 'Bighead') %>%  select(LastAvoid_Calculated)
avoidSilver <- BothExp1_Results %>%  filter(Species == 'Silver') %>%  select(LastAvoid_Calculated)
avoidAvgLarge <- (avoidBighead + avoidSilver)/2
measureAvgLarge <- round(mean(Exp1_Results$Measured))

detach(Exp1_Results)
```

```{r export table for publication, include=FALSE, eval=FALSE}
labels <- c("Tank", "Species", "TA (ppm)", "SE", "Temp (°C)", "SE","Estimated (ppm)","SE","Measured (ppm)","SE")
tab_df(BothExp1_Results,
       title = "Dissolved carbon dioxide avoidance and ambient water parameters",
       col.header = labels,
       footnote = "text",
       show.footnote = FALSE,
       encoding = "Windows-1252",
       digits = 1,
       file = "./Figures/CO2_Avoidance_Table.doc"
    #   file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/CO2 Avoidance Table.doc"
    #  user.viewer = TRUE
)
```

## CO~2~ avoidance figure (boxplots)

```{r Boxplot Figure, message=FALSE, warning=FALSE}
SmallExp1$Tank <- c("Small")
Exp1$Tank <- c("Large")

BothExp1 <- full_join(Exp1, SmallExp1) %>% 
  mutate(Avoid = coalesce(Last_Avoid_Estimated_CO2_ppm,CO2_Output_uAtm))  %>% 
  select(Species,Avoid,Tank,School)

compareTanks <- list(c("Large", "Small"))
compareSpecies <- list(c("Bighead", "Silver"))

CO2_Avoidance_Boxplot_1 <-  ggplot(BothExp1, aes(x = Tank, y = Avoid)) +
  stat_compare_means(label.x = 1.4) +
  geom_boxplot(position = position_dodge(1)) +
  geom_jitter(binaxis = 'y',           # add dots
              width = .3,
               stackdir = 'center',
               size = 3,
               aes(color = Species)) + 
  labs(x = "Tank Size", 
       y = expression("pCO"[2]*" (ppm) at Last Exit")) +
  theme_classic2() + scale_color_grey()

CO2_Avoidance_Boxplot_1 # better compares tanks

CO2_Avoidance_Boxplot_2 <-  ggplot(BothExp1, aes(x = Species, y = Avoid)) +
  stat_compare_means(label.x = 1.4) +
  geom_boxplot(position = position_dodge(1)) +
  geom_jitter(binaxis = 'y',           # add dots
              width = .1,
              stackdir = 'center',
              size = 3,
              aes(color = Tank)) + 
  labs(x = "Species", 
       y = expression("pCO"[2]*" (ppm) at Last Exit")) +
  theme_classic2() + scale_color_discrete(type = c("#7570b3", "#d95f02"))


CO2_Avoidance_Boxplot_2 # better compares species
```

```{r export boxplot fig, include=FALSE, eval=FALSE}
tiff("./Figures/TIF/Fig3_CO2_Avoidance_Boxplot.tif", width = 1800, height = 1200, units = "px", res = 300, compression = c("none"))
CO2_Avoidance_Boxplot_2
dev.off()

ggsave(file = "./Figures/EPS/Fig3_CO2_Avoidance_Boxplot.eps",
       dpi = 300, width = 6, height = 5, units = "in", limitsize = TRUE)
CO2_Avoidance_Boxplot_2
dev.off()
```

## Avoidance Threshold in pH

Calculate the pH at the avoidance threshold using the reverse equation from before: now ppm to pH This code was used to predict the pH to reach during conditioning treatments.

```{r Avoidance Thresholds}
fit <- lm(pH~poly(ppm,2,raw=TRUE), data = curve)

pH <- as.numeric(c("NA"))
ppm <- as.numeric(c(avoidAvgBoth))
threshold <- data.frame(pH, ppm)

threshold$pH <- predict(fit, newdata = threshold)
threshold$avoidAvgRound <- round(avoidAvgBoth, digits = 1)
threshold
```

```{r figure and kruscal test}
modTank <- kruskal.test(Avoid ~ Tank, data = BothExp1)
modTank

modSpecies <- kruskal.test(Avoid ~ Species, data = BothExp1)
modSpecies
```

```{r formatting variables, include = FALSE}
# formatting guidelines: If tens of thousands or more, use a comma. Otherwise, omit.

# paragraph one: both tanks, both species, one avoidance threshold
chi2Tank <- format(round(as.numeric(modTank$statistic), 2), nsmall = 2)
pvalTank <- format(round(as.numeric(modTank$p.value), 2), nsmall = 2)

chi2Species <- format(round(as.numeric(modSpecies$statistic), 2), nsmall = 2)
pvalSpecies <- format(round(as.numeric(modSpecies$p.value), 2), nsmall = 2)

avoidPPM <- format(signif(as.numeric(avoidAvgBoth), 5))

# format for xy.z ppt (parts per thousand)
# avoidPPM <- format(signif(as.numeric(avoidAvgBoth), 3), big.mark = ".", width = 3)
# avoidPPM <- substr(avoidPPM, 1, 4)


pHThreshold <- format(round(as.numeric(threshold$pH), 2), nsmall = 2)

# paragraph two 
smallCO2 <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(LastAvoid_Calculated))), nsmall = 0)
smallCO2_SE <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(LastAvoid_SE))), nsmall = 0)
smallMeasured <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(Measured))))
smallMeasured_SE <- format(round(as.numeric(BothExp1_Results %>% filter(Species == 'Bigheaded') %>% select(Measured_SE))))

# paragraph three
largeBighead <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Bighead') %>% select(LastAvoid_Calculated))))
largeBighead_SE <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Bighead') %>% select(LastAvoid_SE))))
largeSilver <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Silver') %>% select(LastAvoid_Calculated))))
largeSilver_SE <- format(round(as.numeric(BothExp1_Results %>%  filter(Species == 'Silver') %>% select(LastAvoid_SE))))
largeBoth <- format(round(as.numeric(avoidAvgLarge)))
largeMeasured <- format(round(as.numeric(measureAvgLarge)))

largeMeasured_SE <- sd(Exp1$End_Trial_Measured_CO2_ppm, na.rm = TRUE)/sqrt(length(Exp1$End_Trial_Measured_CO2_ppm))
largeMeasured_SE <- format(round(as.numeric(largeMeasured_SE)))
largeMeasured_N <- as.numeric(length(Exp1$End_Trial_Measured_CO2_ppm))
```

# Results

## CO~2~ avoidance threshold

The two chambers experienced relatively little water exchange across the connecting channel due to their circular design and opposing currents.
Water mixed uniformly within each chamber, as evidenced in the large tank by similar pH values in the center of the chamber, immediately next to the exit, and at the gas stone.

There was not a significant difference in dissolved CO~2~ avoidance thresholds between tank size (`r modTank$method`, Χ^2^ = `r chi2Tank`, p = `r pvalTank`) or species (`r modSpecies$method`, Χ^2^ = `r chi2Species`, p = `r pvalSpecies`, Fig 5).
Across all trials in both tanks, bigheaded carp avoided pCO~2~ at approximately `r avoidPPM` ppm.
This threshold pCO~2~ level was used with CO~2~ conditioning treatments by reaching an instantaneous pH of approximately `r pHThreshold`.
Due to tank volume differences, it was possible to reach this threshold within two minutes in the small tank, and ten minutes in the large tank, when addition was stopped for researcher safety.

```{r graph in results, echo=FALSE, message=FALSE}
CO2_Avoidance_Boxplot_2
```

\
**Fig 5. CO~2~ avoidance boxplots.** Bigheaded carps avoided pCO~2~ at `r avoidPPM` ppm on average.
Boxplots include median ± quartiles for each species.
Whiskers extend to 1.5 times IQR.
Dark and light points are from the large and small tank respectively.
Silver and bighead exited the increasing pCO~2~ chamber for the last time at the same threshold (`r modSpecies$method`, Χ^2^ = `r chi2Species`, p = `r pvalSpecies`).\

Avoidance was defined as the last time the individual or the majority of the school (3 or more fish) exited the CO~2~ chamber.
After this last exit, some fish entered the shuttle channel and then turned around without crossing into the increasing pCO~2~ chamber, indicating their avoidance threshold had already been passed.
In the small shuttle tank, schools of five juvenile untrained bighead (n = 5) and silver (n = 5) carp exhibited CO~2~ avoidance at `r smallCO2` ± `r smallCO2_SE` ppm (n = 10, mean ± SE, Table 2) as estimated in CO2Calc.
As a secondary measurement of pCO~2~ from water samples, avoidance occurred at `r smallMeasured` ± `r smallMeasured_SE` ppm (n = 6).

\
In the large shuttle tank, individual bighead carp (n = 10) exhibited CO~2~ avoidance at `r largeBighead` ± `r largeBighead_SE` ppm (Table 1).
Schools of five silver carp (n = 9) last exited the increasing chamber at `r largeSilver` ± `r largeSilver_SE` ppm.
In the large tank, the last exit occurred at `r largeBoth` ppm CO~2~ on average for both species, as estimated by pH (Fig 2).
The average pCO~2~ in the increasing chamber was measured at `r largeMeasured` ± `r largeMeasured_SE` ppm (n = `r largeMeasured_N`) three minutes after the last exit.

**Table 1. Carbon dioxide avoidance and ambient water parameters in both tanks.** The pCO~2~ is estimated from instantaneous pH at the last exit and measured from water samples.

```{r table in results, echo=FALSE, message=FALSE}
CO2_Avoidance_Table
```

\
\
\

## Unused code below, cut before final archiving

```{r Boxplot Figure OPTIONS, include=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
SmallExp1$Tank <- c("Small")
Exp1$Tank <- c("Large")

SmallExp1$School <- c("School")
Exp1$School <- ifelse(Exp1$Species == 'Bighead', "Individual", "School")

BothExp1 <- full_join(Exp1, SmallExp1) %>% 
  mutate(Avoid = coalesce(Last_Avoid_Estimated_CO2_ppm,CO2_Output_uAtm))  %>% 
  select(Species,Avoid,Tank,School)

CO2_Avoidance_Boxplot_1 <-  ggplot(BothExp1, aes(x = Tank, y = Avoid)) +
  stat_compare_means(label.x = 1.4) +
  geom_boxplot(position = position_dodge(1)) +
  geom_point(binaxis = 'y',           # add dots
               stackdir = 'center',
               position = position_dodge(.15),
               size = 3,
               aes(color = Species, shape = School)) + 
  labs(x = "Tank Size", 
       y = expression("pCO"[2]*" (ppm) at Last Exit")) +
  theme_classic2() + scale_color_grey()

CO2_Avoidance_Boxplot_1 # better compares tanks

CO2_Avoidance_Boxplot_2 <-  ggplot(BothExp1, aes(x = Species, y = Avoid)) +
  stat_compare_means(label.x = 1.4) +
  geom_boxplot(position = position_dodge(1)) +
  geom_point(binaxis = 'y',           # add dots
               stackdir = 'center',
               position = position_dodge(.15),
               size = 3,
               aes(color = Tank, shape = School)) + 
  labs(x = "Species", 
       y = expression("pCO"[2]*" (ppm) at Last Exit")) +
  theme_classic2() + scale_color_grey()

CO2_Avoidance_Boxplot_2 # better compares species
```
