---
title: "CO2_Conditioning"
author: "Jackie Culotta"
date: "2022-11-15"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
This Rmd file summarizes water parameters for the conditioning treatments in both tanks.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library('dplyr')
library('ggpubr')
library('rstatix')

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()
```

```{r load data, message=FALSE, warning=FALSE}
# Import csv files
largeTank <- read.csv("Conditioning_LargeTank.csv")
smallTank <- read.csv("Conditioning_SmallTank.csv")

# Change columns to factors
factorColumns <- c("Fish_ID", "Conditioning", "Speaker", "Species", "Day")
largeTank <- largeTank %>% 
  mutate_at(factorColumns, factor) %>% 
  mutate_if(is.character, as.numeric)

str(largeTank)

factorColumns <- c("School_ID", "Conditioning", "Speaker", "Species")
smallTank <- smallTank %>% 
  mutate_at(factorColumns, factor) %>% 
  mutate_if(is.character, as.numeric)

str(smallTank)

rm(factorColumns)
```

```{r summary stats}
# Remove rows with NAs
smallTank2 <- smallTank %>% filter(!is.na(Mid_ppm) | !is.na(End_ppm))
largeTank2 <- largeTank %>% filter(!is.na(Start_ppm) | !is.na(End_ppm))

# summary in small tank
smallSummary <- smallTank2 %>%
group_by(Conditioning) %>%
  summarize(
    smallMid_ppm = mean(Mid_ppm),
    smallMid_SE = sd(Mid_ppm)/sqrt(length(Mid_ppm)),
    smallEnd_ppm = mean(End_ppm),
    smallEnd_SE = sd(End_ppm)/sqrt(length(End_ppm))
    )
smallSummary <- as.data.frame(smallSummary)
smallSummary

# summary in large tank
largeSummary <- largeTank2 %>%
group_by(Conditioning) %>%
  summarize(
    largeStart_ppm = mean(Start_ppm),
    largeStart_SE = sd(Start_ppm)/sqrt(length(Start_ppm)),
    largeEnd_ppm = mean(End_ppm),
    largeEnd_SE = sd(End_ppm)/sqrt(length(End_ppm))
    )
largeSummary <- as.data.frame(largeSummary)
largeSummary
```
```{r in text, include=FALSE}
# small tank midpoint
smallMid_ppm <- smallSummary %>% slice(1) %>% select(smallMid_ppm)
smallMid_ppm <- format(signif(as.numeric(smallMid_ppm), 3))

smallMid_SE <- smallSummary %>% slice(1) %>% select(smallMid_SE)
smallMid_SE <- format(signif(as.numeric(smallMid_SE), 3))

# small tank end point

smallEnd_ppm <- smallSummary %>% slice(1) %>% select(smallEnd_ppm)
smallEnd_ppm <- format(signif(as.numeric(smallEnd_ppm), 3))

smallEnd_SE <- smallSummary %>% slice(1) %>% select(smallEnd_SE)
smallEnd_SE <- format(signif(as.numeric(smallEnd_SE), 3))


# large tank start
CO2largeStart_ppm <- largeSummary %>% slice(2) %>% select(largeStart_ppm)
CO2largeStart_ppm <- format(signif(as.numeric(CO2largeStart_ppm), 3))

CO2largeStart_SE <- largeSummary %>% slice(2) %>% select(largeStart_SE)
CO2largeStart_SE <- format(signif(as.numeric(CO2largeStart_SE), 3))


AIRlargeStart_ppm <- largeSummary %>% slice(1) %>% select(largeStart_ppm)
AIRlargeStart_ppm <- format(signif(as.numeric(AIRlargeStart_ppm), 3))

AIRlargeStart_SE <- largeSummary %>% slice(1) %>% select(largeStart_SE)
AIRlargeStart_SE <- format(signif(as.numeric(AIRlargeStart_SE), 3))


# large tank end point
CO2largeEnd_ppm <- largeSummary %>% slice(2) %>% select(largeEnd_ppm)
CO2largeEnd_ppm <- format(signif(as.numeric(CO2largeEnd_ppm), 3))

CO2largeEnd_SE <- largeSummary %>% slice(2) %>% select(largeEnd_SE)
CO2largeEnd_SE <- format(signif(as.numeric(CO2largeEnd_SE), 3))

AIRlargeEnd_ppm <- largeSummary %>% slice(1) %>% select(largeEnd_ppm)
AIRlargeEnd_ppm <- format(signif(as.numeric(AIRlargeEnd_ppm), 3))

AIRlargeEnd_SE <- largeSummary %>% slice(1) %>% select(largeEnd_SE)
AIRlargeEnd_SE <- format(signif(as.numeric(AIRlargeEnd_SE), 3))
```

# Results

In the small tank, the pCO~2~ halfway into the conditioning treatment was `r smallMid_ppm` ± `r smallMid_SE` ppm. Dissolved CO~2~ reached `r smallEnd_ppm` ± `r smallEnd_SE` ppm at the end of the treatment.
\
In the large tank, dissolved CO~2~ before conditioning was `r CO2largeStart_ppm` ± `r CO2largeStart_SE` ppm for schools trained with CO~2~, and `r AIRlargeStart_ppm` ± `r AIRlargeStart_SE` ppm for schools trained with air. 
At the end of the conditioning treatment, the dissolved CO~2~ for carp conditioned with air was `r AIRlargeEnd_ppm` ± `r AIRlargeEnd_SE` ppm. 
Carp in the CO~2~ treatment (N = 10) were conditioned at `r CO2largeEnd_ppm` ± `r CO2largeEnd_SE` ppm by the conditioning period endpoint.  

