---
title: "CO2_Conditioning"
author: "Jackie Culotta"
date: "2022-11-15"
output:
  html_document: default
  word_document: default
---
This Rmd file summarizes water parameters for the conditioning treatments in both tanks.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library('dplyr')
library('ggpubr')
library('rstatix')

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()
```

```{r load data}
# Import csv files
largeTank <- read.csv("Conditioning_LargeTank.csv")
smallTank <- read.csv("Conditioning_SmallTank.csv")

# Change columns to factors
factorColumns <- c("Fish_ID", "Conditioning", "Speaker", "Species", "Day")
largeTank <- largeTank %>% 
  mutate_at(factorColumns, factor) %>% 
  mutate_if(is.character, as.numeric)

str(largeTank)

factorColumns <- c("School_ID", "Conditioning", "Speaker", "Species")
smallTank <- smallTank %>% 
  mutate_at(factorColumns, factor) %>% 
  mutate_if(is.character, as.numeric)

str(smallTank)
```

```{r summary stats}
# Remove rows with NAs
smallTank2 <- smallTank %>% filter(!is.na(Mid_ppm) | !is.na(End_ppm))
largeTank2 <- largeTank %>% filter(!is.na(Start_ppm) & !is.na(End_ppm))

# summary in small tank
smallSummary <- smallTank2 %>%
group_by(Conditioning) %>%
  summarize(
    smallMid_ppm = mean(Mid_ppm),
    smallEnd_ppm = mean(End_ppm))
smallSummary <- as.data.frame(smallSummary)
smallSummary

# summary in large tank
largeSummary <- largeTank2 %>%
group_by(Conditioning) %>%
  summarize(
    largeStart_ppm = mean(Start_ppm),
    largeEnd_ppm = mean(End_ppm))
largeSummary <- as.data.frame(largeSummary)
largeSummary
```
```{r in text, include=FALSE}
# CHECK BACK on how Al wants these formatted !

largeEnd_ppm <- largeSummary %>% slice(2) %>% select(largeEnd_ppm)
largeEnd_ppm <- format(signif(as.numeric(largeEnd_ppm), 3), big.mark = ",")
```

# Results
## Small Tank

pCO~2~ at conditioning midpoint was `r smallMidpoint` ± `r smallMidpoint_SE` ppm. 
pCO~2~ at conditoining endpoint was `r smallEnd`± `r smallEnd_SE` ppm. 

## Large tank
In the large tank, dissolved CO~2~ before trial initiation was 2310 ± 239 ppm for schools trained with CO~2~, and 1459 ± 45 ppm for schools trained with air. 
At the trial endpoint, the dissolved CO2 for carp undergoing conditioning with air was 1494 ± 65 ppm. 
Carp in the CO~2~ treatment (N = 10) were conditioned at 34577 ± 2,068 by the conditioning period endpoint.  