---
title: "SmallTank"
author: "Jackie Culotta"
date: "2022-06-14"
output:
  html_document: default
  word_document: default
---

This Rmd file analyzes the time until first exit and first return for bigheaded carp in the small tank.
Mixed ANOVAs and t-tests are computed.
Boxplots for the associated manuscript are generated.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# Set Up & Data Management

# Load Required Packages
library('dplyr')
library('ggplot2')
library('ggpubr')
library('tidyverse')
library('rstatix')
library('lme4') # mixed ANOVAs

data <- read.csv("SmallTank_Phonotaxis.csv") %>% 
  filter(VID_ID >= 53) %>% # first 52 trials used pure tones, all later trials used broadband
  filter(Days_Since_Train < 29) %>% # all broadband trials went at least 29 days
  select(-c(Stopped_Sound_On_Return, Exit_N, Return_N)) %>% 
  rename(
    Sound = Frequency_Hz, # whether BB sound was on (BB) or off (acclimation)
  )


# Convert variables to factors
data$School_ID <- as.factor(data$School_ID) 
data$Speaker <- as.factor(data$Speaker)
data$Sound <- as.factor(data$Sound)
data$Trained <- as.factor(data$Trained)
data$Species <- as.factor(data$Species)
data$Day <- as.factor(data$Days_Since_Train) # Day needs to be factor for Boxplots
str(data) # confirm variable types

# Convert other variable types
data$Exit_s <- as.integer(data$Exit_s)
data$Return_s <- as.integer(data$Return_s)
data$Angle <- as.integer(data$Angle)
data$Temp <- as.numeric(data$Temp)

data$logQuiet <- log(data$Return_s)
data$logExit <- log(data$Exit_s)

str(data)
summary(data)
```

Assumption checking for Mixed ANOVA (e.g. repeated measures) from <https://www.datanovia.com/en/lessons/mixed-anova-in-r/>

# First Exit

```{r first exit anova assumptions}
data2 <- data %>% filter(!is.na(Exit_s), !is.infinite(Exit_s))

# identify outliers 
outliers <- data2 %>% 
  group_by(Day, Trained, Sound) %>%
  identify_outliers(Exit_s) # beyond 1.5 x IQR

dataNoOut <- data2 %>% anti_join(outliers)

# normality assumption
ggqqplot(dataNoOut, "Exit_s", ggtheme = theme_bw()) +
  facet_grid(Day ~ Trained) 
# if points fall along expected line in each grid, meets assumption

# Homogeneity of variances assumption
dataNoOut %>%
  group_by(Day) %>%
  levene_test(Exit_s ~ Trained)
# if p > .05 for each day assumption is met

# Homogeneity of covariances assumption
box_m(dataNoOut[, "Exit_s", drop = FALSE], dataNoOut$Trained)
# if p is > 0.001 this assumption is met
# Note: Box's M test is very sensitive
```

```{r first exit anova computation}
# mixed ANOVA computation
mod1 <- lmer(Exit_s ~ Trained*Sound + Days_Since_Train + (1|School_ID), data = dataNoOut)
Anova(mod1)

mod2 <- lmer(Exit_s ~ Trained*Sound*Days_Since_Train + (1|School_ID), data = dataNoOut)
Anova(mod2)

anova(mod1,mod2) # if p < 0.05, reject the simpler model
mod <- mod2 # better model of the two is THE model
```

```{r SmallBoxplotExit figure}
# Including only sound trials, ie excluding acclimation period
data3 <- dataNoOut %>% filter(Sound == 'BB')

SmallboxplotExit <- ggplot(data3, aes(x = Day,y = Exit_s, fill = Trained)) +
  geom_boxplot(position = position_dodge(1)) +
  stat_compare_means(aes(group = Trained), method = "t.test", 
                     label = "p.signif", hide.ns = TRUE) + 
  labs(x = "Day", y ="Seconds to Exit Sound Chamber") +
  scale_fill_brewer(palette = "Greys") +
  theme_classic2()
  # geom_dotplot(binaxis='y', stackdir='center', position=position_dodge(1)) # add dots

 # SmallboxplotExit
```

```{r day 1 t test}
# T test for day 1
data4 <- data3 %>% filter(Day == 1)
modDay1 <- t.test(data = data4, Exit_s ~ Trained)
```

```{r in text numbers, include=FALSE}
# ANOVA (including acclimation)
modSummary <- as.data.frame(Anova(mod)) %>% rename(pval = 3)

soundPval <- modSummary %>% slice(2) %>% select(pval)
soundPval <- format(round(as.numeric(soundPval), 3), nsmall = 2, big.mark = ",")

soundChisq <- modSummary %>% slice(2) %>% select(Chisq)
soundChisq <- format(round(as.numeric(soundChisq), 2), nsmall = 2, big.mark = ",")


# t test, day 1 sound only
day1Untrained <- modDay1$estimate[1]
day1Untrained <- format(round(as.numeric(day1Untrained), 1), nsmall = 1, big.mark = ",")

day1Trained <- modDay1$estimate[2]
day1Trained <- format(round(as.numeric(day1Trained), 1), nsmall = 1, big.mark = ",")

day1Method <- modDay1$method

day1Stat <- modDay1$statistic
day1Stat <- format(round(as.numeric(day1Stat), 2), nsmall = 2, big.mark = ",")

day1DF <- modDay1$parameter
day1DF <- format(round(as.numeric(day1Stat), 2), nsmall = 2, big.mark = ",")

day1Pval <- modDay1$p.value
day1Pval <- format(round(as.numeric(day1Pval), 2), nsmall = 2, big.mark = ",")
```

## Results

All trained carp exhibited negative phonotaxis by shuttling out of the sound chamber.
Two untrained schools did not exit the sound chamber within the allotted five minutes of acoustic playback.
Sound was the only predictor that had a significant effect on seconds until first exit (Mixed ANOVA, Χ^2^ = `r soundChisq`, p = `r soundPval`).
Notably on the first day of phonotaxis trials, trained schools initially left the sound chamber in `r day1Trained` s, whereas untrained schools took `r day1Untrained` s to exit (`r day1Method`, df = `r day1DF`, t = `r day1Stat`, p = `r day1Pval`).
There were no differences in time to exit due to training treatment on later sound presentations (Fig SmallBoxplotExit).

```{r Small boxplot exit figure}
SmallboxplotExit
```

**Fig SmallBoxplotExit. Seconds in the sound chamber until exiting to the quiet chamber.** On the first day, trained schools initially left the sound chamber in `r day1Trained` s, whereas untrained schools took `r day1Untrained` s to exit (`r day1Method`, df = `r day1DF`, t = `r day1Stat`, p = `r day1Pval`).
There were no differences in time to exit due to training treatment on later sound presentations.

```{r export exit boxplot, eval=FALSE, include=FALSE}
ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig6_smallboxplotExit.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
SmallboxplotExit
dev.off()

tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig6_smallboxplotExit.tif", width = 2100, height = 1200, units = "px", res = 300, compression = c("none"))
 SmallboxplotExit
dev.off()
```

```{r cleanup, include=FALSE}
rm(mod)
rm(mod2)
rm(mod3)
rm(mod4)
rm(modDay1)
rm(SmallboxplotExit)
rm(outliers)
rm(dataNoOut)
rm(data4)
rm(data3)
rm(data2)
```

# Time to Return

```{r first return anova assumptions, echo=TRUE}
data2 <- data %>% filter(!is.na(Return_s), !is.infinite(Return_s))

# identify outliers (both species)
outliers <- data2 %>% 
  group_by(Days_Since_Train, Trained, Sound) %>%
  identify_outliers(Return_s)

dataNoOut <- data2 %>% anti_join(outliers) # remove outliers

# normality assumption
ggqqplot(dataNoOut, "Return_s", ggtheme = theme_bw()) +
  facet_grid(Day ~ Trained) 
# points fall along expected line in each grid, meets assumption

# Homogeneity of variances assumption
dataNoOut %>%
  group_by(Day) %>%
  levene_test(Return_s ~ Trained)
# if p > .05 for each day assumption is met

# Homogeneity of covariances assumption
box_m(dataNoOut[, "Return_s", drop = FALSE], dataNoOut$Trained)
# if p is > 0.001 this assumption is met
# Note: Box's M test is very sensitive
```

```{r return anova computation}
# mixed ANOVA computation
mod1 <- lmer(Return_s ~ Trained*Sound + Days_Since_Train + (1|School_ID), data = dataNoOut)

mod2 <- lmer(Return_s ~ Trained*Sound*Days_Since_Train + (1|School_ID), data = dataNoOut)

anova(mod1,mod2)# if p < 0.05, reject simpler model
mod <- mod2 # better model wins

Anova(mod)
```

```{r create return boxplot}
# Including only sound trials, ie excluding acclimation period
data3 <- dataNoOut %>% filter(Sound == 'BB')

smallboxplotReturn <- ggplot(data3, aes(x = Day,y=Return_s, fill = Trained)) +
  geom_boxplot(position = position_dodge(1)) +
  stat_compare_means(aes(group = Trained), method = "t.test", 
                     label = "p.signif", hide.ns = TRUE) + 
  labs(x = "Day", y = "Seconds in Quiet Chamber until Return") + 
  scale_fill_brewer(palette = "Greys") +
  theme_classic2()
  # geom_dotplot(binaxis='y', stackdir='center', position=position_dodge(1)) # add dots

# smallboxplotReturn
```

```{r export return boxplot, eval=FALSE, include=FALSE}
ggsave(file = "C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/EPS/Fig8_smallboxplotReturn.eps",
       dpi = 300, width = 5, height = 5, units = "in", limitsize = TRUE)
smallboxplotReturn
dev.off()

tiff("C:/Users/jacki/OneDrive/Documents (Onedrive)/Thesis/Analysis/R_Statistics/Figures/TIF/Fig8_smallboxplotReturn.tif", width = 2100, height = 1200, units = "px", res = 300, compression = c("none"))
 smallboxplotReturn
dev.off()
```

```{r t tests day 1 and 7}
# T test for day 1
data4 <- data3 %>% filter(Day == 1)
modDay1 <- t.test(data = data4, Return_s ~ Trained)
modDay1

# T test for day 7
data4 <- data3 %>% filter(Day == 7)
modDay7 <- t.test(data = data4, Return_s ~ Trained)
modDay7
```

## Results

```{r anova in text numbers, include=FALSE}
modSummary <- as.data.frame(Anova(mod)) %>% rename(pval = 3)

# training
trainPval <- modSummary %>% slice(1) %>% select(pval)
trainPval <- format(round(as.numeric(trainPval), 3), nsmall = 2, big.mark = ",")
trainChisq <- modSummary %>% slice(1) %>% select(Chisq)
trainChisq <- format(round(as.numeric(trainChisq), 2), nsmall = 2, big.mark = ",")

# sound
soundPval <- modSummary %>% slice(2) %>% select(pval)
soundPval <- format(round(as.numeric(soundPval), 3), nsmall = 2, big.mark = ",")
soundChisq <- modSummary %>% slice(2) %>% select(Chisq)
soundChisq <- format(round(as.numeric(soundChisq), 2), nsmall = 2, big.mark = ",")

# day
dayPval <- modSummary %>% slice(3) %>% select(pval)
dayPval <- format(round(as.numeric(dayPval), 3), nsmall = 2, big.mark = ",")
dayChisq <- modSummary %>% slice(3) %>% select(Chisq)
dayChisq <- format(round(as.numeric(dayChisq), 2), nsmall = 2, big.mark = ",")

# interaction
interactPval <- modSummary %>% slice(7) %>% select(pval)
interactPval <- format(round(as.numeric(interactPval), 3), nsmall = 2, big.mark = ",")
interactChisq <- modSummary %>% slice(7) %>% select(Chisq)
interactChisq <- format(round(as.numeric(interactChisq), 2), nsmall = 2, big.mark = ",")
```

The assumptions of no outliers, normality, and homogeneity of variances are met.
A significant result for Box's M-test for Homogeneity of Covariance Matrices is likely due to unequal sample sizes between trained and untrained groups after removal of outliers.
A mixed ANOVA is run noting this violation.

Training with CO~2~ had a marginally significant effect on seconds to return to the sound chamber (Mixed ANOVA, Χ^2^ = `r trainChisq`, p = `r trainPval`).
Presence of the acoustic signal had a highly significant effect on time to return (Χ^2^ = `r soundChisq`, p = `r soundPval`).
Additionally, the number of days since training had a significant effect on time to return (Χ^2^ = `r dayChisq`, p = `r dayPval`).
Notably, there was a significant three-way interaction between training, presence of sound, and the number of days since trial onset (Χ^2^ = `r interactChisq`, p = `r interactPval`).

```{r t test in text numbers, include=FALSE}
# t test, day 1 sound only
day1Untrained <- modDay1$estimate[1]
day1Untrained <- format(round(as.numeric(day1Untrained), 1), nsmall = 1, big.mark = ",")

day1Trained <- modDay1$estimate[2]
day1Trained <- format(round(as.numeric(day1Trained), 1), nsmall = 1, big.mark = ",")

day1Method <- modDay1$method

day1Stat <- modDay1$statistic
day1Stat <- format(round(as.numeric(day1Stat), 2), nsmall = 2, big.mark = ",")

day1DF <- modDay1$parameter
day1DF <- format(round(as.numeric(day1Stat), 2), nsmall = 2, big.mark = ",")

day1Pval <- modDay1$p.value
day1Pval <- format(round(as.numeric(day1Pval), 2), nsmall = 2, big.mark = ",")

# day 7
day7Untrained <- modDay7$estimate[1]
day7Untrained <- format(round(as.numeric(day7Untrained), 1), nsmall = 1, big.mark = ",")

day7Trained <- modDay7$estimate[2]
day7Trained <- format(round(as.numeric(day7Trained), 1), nsmall = 1, big.mark = ",")

day7Method <- modDay7$method

day7Stat <- modDay7$statistic
day7Stat <- format(round(as.numeric(day7Stat), 2), nsmall = 2, big.mark = ",")

day7DF <- modDay7$parameter
day7DF <- format(round(as.numeric(day7Stat), 2), nsmall = 2, big.mark = ",")

day7Pval <- modDay7$p.value
day7Pval <- format(round(as.numeric(day7Pval), 2), nsmall = 2, big.mark = ",")
```

On the first day of phonotaxis trials, trained schools spent `r day1Trained` s in the quiet chamber before returning to the loud chamber, whereas untrained schools took only `r day1Untrained` s to return (`r day1Method`, df = `r day1DF`, t = `r day1Stat`, p = `r day1Pval`).
One week after training, trained schools spent `r day7Trained` s in the quiet chamber before returning, compared to `r day7Untrained` s for untrained schools (`r day7Method`, df = `r day7DF`, t = `r day7Stat`, p = `r day7Pval`).
There were no differences in time to return to the sound chamber due to training treatment on sound presentations after one week (Fig SmallBoxplotReturn).

```{r print return boxplot}
smallboxplotReturn
```

**Fig smallboxplotReturn Seconds in the quiet chamber until returning to the sound chamber.** On the first day, trained schools stayed in the quiet chamber for `r day1Trained` s, whereas untrained schools returned after only `r day1Untrained` s (`r day1Method`, df = `r day1DF`, t = `r day1Stat`, p = `r day1Pval`).
On the seventh day, trained schools returned after `r day7Trained` s, compared to `r day7Untrained` s for untrained schools (`r day7Method`, df = `r day7DF`, t = `r day7Stat`, p = `r day7Pval`).
There were no differences in time to return due to training treatment beyond one week after conditioning.

```{r habituation}
# Trained
data4 <- data3 %>% filter(Trained == '2 days') %>% filter(Day == 1 | Day == 4)
modTrained <- t.test(data = data4, Return_s ~ Day)
modTrained

data4 <- data3 %>% filter(Trained == '0 days') %>% filter(Day == 1 | Day == 4)
moduntrained <- t.test(data = data4, Return_s ~ Day)
moduntrained
```

```{r habituation t tests, include=FALSE}
# Trained

day1Trained <- modTrained$estimate[1]
day4Trained <- modTrained$estimate[2]
TrainedDiff <- day1Trained - day4Trained
TrainedDiff <- format(round(as.numeric(TrainedDiff), 1), nsmall = 1, big.mark = ",")

TrainedMethod <- modTrained$method

TrainedStat <- modTrained$statistic
TrainedStat <- format(round(as.numeric(TrainedStat), 2), nsmall = 2, big.mark = ",")

TrainedDF <- modTrained$parameter
TrainedDF <- format(round(as.numeric(TrainedDF), 2), nsmall = 2, big.mark = ",")

TrainedPval <- modTrained$p.value
TrainedPval <- format(round(as.numeric(TrainedPval), 2), nsmall = 2, big.mark = ",")

# Untrained
day1untrained <- moduntrained$estimate[1]
day4untrained <- moduntrained$estimate[2]
untrainedDiff <- day1untrained - day4untrained
untrainedDiff <- format(round(as.numeric(untrainedDiff), 1), nsmall = 1, big.mark = ",")

untrainedMethod <- moduntrained$method

untrainedStat <- moduntrained$statistic
untrainedStat <- format(round(as.numeric(untrainedStat), 2), nsmall = 2, big.mark = ",")

untrainedDF <- moduntrained$parameter
untrainedDF <- format(round(as.numeric(untrainedDF), 2), nsmall = 2, big.mark = ",")

untrainedPval <- moduntrained$p.value
untrainedPval <- format(round(as.numeric(untrainedPval), 2), nsmall = 2, big.mark = ",")
```

Untrained fish returned to the sound chamber `r untrainedDiff` s faster on the second sound presentation, which occurred on the 4th day after conditioning, than they had during the first presentation, although this difference was not significant (df = `r untrainedDF`, t = `r untrainedStat`, p = `r untrainedPval`).
In contrast, trained fish returned to the sound chamber `r TrainedDiff` s faster than they had for the second sound presentation (df = `r TrainedDF`, t = `r TrainedStat`, p = `r TrainedPval`).
This suggests trained fish may have habituated to the sound stimulus on the second presentation in a way the untrained fish did not.

# Extra unused code included in source file

```{r misc, include=FALSE, eval=FALSE, echo=FALSE}
# Replace acronyms with full text
data$Species <- gsub("BH", "Bighead", data$Species)
data$Species <- gsub("S", "Silver", data$Species)

data$Speaker <- gsub("R", "Right", data$Speaker)
data$Speaker <- gsub("L", "Left", data$Speaker)
```

```{r scatterplot1, include=FALSE, eval=FALSE, echo=FALSE}
# Scatterplot linear model for first return
p <- ggplot(dataNoOut, aes(x=Days_Since_Train,y=Return_s, color=Trained)) + geom_point() +
  geom_smooth(method="lm", se=TRUE, level=0.95, fullrange=FALSE, aes(fill=Trained)) +
  # labs(x="Days Since Conditioning", y="Seconds in Opposite Chamber till First Return") 
  coord_cartesian(xlim = c(1, 28), ylim = c(0,300)) + # crop display area
  scale_y_continuous(name = "Seconds till Return to Sound Chamber") +
  scale_x_continuous(name = "Days", breaks = seq(1,28,3)) +
  ggtitle("Silver and Bighead Carp")
p + theme_classic2()
```

```{r summary stats method 2, include=FALSE, eval=FALSE}
# Another way to pull summary stats for each group
dataSummary <- data3 %>% # or dataNoOut to include acclimation
group_by(Sound,Trained,Day) %>%
  summarize(
    Quiet_s = mean(Return_s),
    Quiet_s_sd = sd(Return_s))
dataSummary <- as.data.frame(dataSummary)
dataSummary

day1Trained <- dataSummary %>% slice(8) %>% select(Quiet_s)
day4Trained <- dataSummary %>% slice(9) %>% select(Quiet_s)
day1To4Trained <- day1Trained - day4Trained
day1To4Trained <- format(round(as.numeric(day1To4Trained), 1), nsmall = 1, big.mark = ",")
day1To4Trained
```
